Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-15 paper.tex

However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material.
Others state that an ARG is equivalent to the sequence of marginal
% Also SARGE paper
trees along the genome~\citep{mirzaei2016rent}

Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/gc-perf}
\end{center}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/mutation-perf}
\caption{\label{fig-mutation-perf} Time required to run
\texttt{sim\_mutations} on tree sequences generated
by \texttt{sim\_ancestry} (with a population size of $10^4$
and recombination rate of $10^{-8}$) for varying (haploid) sample
size and sequence length. We run 10 replicate mutation simulations
each for three different mutation rates, and report the average
CPU time required (Intel Core i7-9700).
(A) Holding sequence length fixed at 10 megabases and varying the
number of samples (tree tips) from 10 to 100,000.
(B) Holding number of samples fixed at 1000, and varying the sequence
length from 1 to 100 megabases.}
\end{figure}

@@

Fig.~\ref{fig-mutation-perf} shows that the mutation generation
machinery in \msprime\ is extremely efficient, requiring
fractions of a second to generate mutations for most simulations.
For example, the longest running simulation in
Fig.~\ref{fig-mutation-perf}B required less than 2 seconds to
generate an average of 1.5 million mutations over 137,081 trees
in a tree sequence with 508,125 edges.
Generating mutations for a tree sequence is particularly efficient
because we can work edge-by-edge rather than having to
consider each tree independently.

\jkcomment{We probably want to keep a version of this narrative too.}
For example, we simulated mutations under the JC69
model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim 280000$ mutations) in less that 1.5 seconds, including
the time required for file input and output. While a systematic
comparison of run-times is not feasible here, this is orders
of magnitude faster than methods such as
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}
and at least competitive with the recent method from
\cite{demaio2021phastsim}.

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
% also SARGE
opens the possibility for new applications of coalescent simulations.
With lots of applications, e.g., [cite Osmond and Coop].
Also \stdpopsim\ is cool~\citep{adrion2020community}

>>>>

\jkcomment{If we're going to have this paragraph of ARG bashing then
I guess we'll have to do more than just single out Iain and Aylwyn.}
However, this original definition of the ARG in terms of a stochastic process
is rarely what is meant when the term is currently used. The term has come to
essentially mean any representation of genetic ancestry in the presence
of recombination as a graph-like structure; for example,
\cite{mathieson2020ancestry} informally define an ARG as the subset of an
individual's pedigree from which they have inherited ancestral material. Such
imprecision is unfortunate, since there are any number of graphical encodings of
genetic ancestry, and referring to the original Griffiths papers does little to
clarify. As we see below, there are situations in which the precise information
stored in the graph is important, and conflating a high-level
discussion of genetic ancestry with this mathematical object is not
helpful. The point of technical terms is to provide clarity and
precision on what is being discussed, and it is now essentially
impossible to know what is meant by the term ``ARG'';
we therefore use the term ``genetic ancestry'' in high-level
discussions.

@@

\begin{figure}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf_log.pdf}
% \includegraphics[width = \textwidth]{figures/gene_conversion/gc_perf.pdf}
\caption{\label{fig-gc-perf}Comparison of the running time
for E-coli simulations using \msprime\ and SimBac for varying
sample sizes. The parameters used are chosen to mimic the simulation
of \textit{E. coli} genomes. The genome length was set to 4.5 mb, the scaled per site
gene conversion rate was set to $N_e g = \gamma = 0.015$ and
the mean gene conversion tract length was set to $L = 500$, which is close to
estimated parameters for \textit{E.\ coli}, i.e.\ an effective populations size
$N_e = 1.8e^8$ and a per site per generation conversion rate $g = 8.9e^{-11}$
with mean tract length $L = 542$~\citep{lapierre2016the}.}
\end{figure}

@@

@@

The mutation generation machinery in \msprime\ is extremely
efficient. For example, we simulated mutations under the JC69
model for a tree with $10^6$ leaves over $10^4$ sites (resulting
in $\sim 280000$ mutations) in less that 1.5 seconds, including
the time required for file input and output. While a systematic
comparison of run-times is not feasible here, this is orders
of magnitude faster than methods such as
Seq-Gen~\citep{rambaut1997seq} and
Pyvolve~\citep{spielman2015pyvolve}
and at least competitive with the recent method from
\cite{demaio2021phastsim}.

@@

% Finish up with a paragraph saying that there loads of exciting
% new things that simulation can do. We need MORE sims!!!
The ability to infer trees at scale for the first time
in the presence of
recombination~\citep{harris2019database,kelleher2019inferring,
speidel2019method,tang2019genealogy}
opens the possibility for new applications of coalescent simulations.
Also \stdpopsim\ is cool~\citep{adrion2020community}
